---
title: "Waller SISMID 2024:  Houston Crime Slippery Slopes"
author: "Lance A. Waller"
date: "7/17/2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Houston Crime Geographically Weighted Regression Example

This file reads in data from a shapefile of Houston census tracts (using the same code in the markdown file we used for Lecture 2). 

Next we will use "spgwr" to fit geographically weighted regression to see if the associations between tract-level (standardized) alcohol sales or illegal drug arrests are correlated with tract-level violent crime rates.

***

***Same code as in GIS example to open libraries, set working directory, and read in the Houston shapefile.***

```{r, output=F, message=F, warning=F}
##Load libraries

pacman::p_load(sf, #replaces "maptools", "rgdal" and other deprecated packages 
               tmap, #helps with plotting your map
               RColorBrewer, # creates nice color schemes
               ClassInt,  # finds class intervals for continuous variables
               GWmodel, # Adds the geographically weighted regression functions
               tidyverse,
               here # For constructing filepaths relative to root directory
               )


##Read in shapefile - Houston Census Tracts
houston = read_sf(dsn = here("data") , layer = "HoustonENAR2012final")

##Map the tracts using plot...
plot(houston['POP2000'],main="Population 2000",
      sub="Quantile (Equal-Frequency) Class Intervals")

```


***

That's a pretty vivid color scheme, let's use tmap's tm_shape to draw nicer maps of the (standardized) violence rate, and the two covariates.  (Similar code to that from Lecture 2).

***


=======
# The data table has a lot of census data and various transformations
# of the violent crime, alcohol sales, and drug arrest data.  The next
# section pulls the values we want.

# Outcome:  Number of violent crimes by tract
violence = houston$violence_2

# Divide by the 2000 population to get the rate
violence.rate = violence/houston$tot_pop

# Covariate 1 (log standardized total alcohol sales)
Z.log.total = houston$Zl_total

# Covariate 2 (log standardized illegal drug arrests)
Z.log.drug = houston$Zl_drug

------------------------------------------------------------------------



```{r}
# Divide by the 2000 population to get the rate
houston$violence.rate = houston$violence_2/houston$tot_pop

# Look at all 9 hex codes in BuPu
#scales::show_col(brewer.pal(9,name="BuPu"))
# Saves all 9 hex codes to a palette
all_colors <- brewer.pal(9, "BuPu")
# Subset to start from the third color
subset_colors <- all_colors[3:(1+6)]
#Display the colors
#subset_colors

violence_map <- tm_shape(houston) + 
  tm_fill('violence.rate', 
style='quantile', 
#midpoint = median(houston$violence.rate),
palette=subset_colors, 
title='Violent Crimes \n Per Capita') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Violent Crime Rate in Houston, TX",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=1.2, legend.title.size=0.5,  
            legend.text.size=0.5)
violence_map
```


------------------------------------------------------------------------

Next, map standardized log total alcohol sales. 

```{r}
alc_map <- tm_shape(houston) + 
  tm_fill('Zl_total', 
style='quantile', 
palette=subset_colors, 
midpoint = mean(houston$Zl_total),
title='Standardized Log \n Total Alcohol Sales') + # "\n" moves text to the next line 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Alcohol Sales in Houston, TX",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=1.2, legend.title.size=0.5,  
            legend.text.size=0.5)
alc_map
```

------------------------------------------------------------------------

Finally, plot standardized log illegal drug arrests. 

```{r}
drug_map <- tm_shape(houston) + 
  tm_fill('Zl_drug', 
style='quantile', 
palette=subset_colors, 
midpoint=mean(houston$Zl_drug),
title='Standardized Log \n Illegal Drug Arrests') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Illegal Drug Arrests in Houston, TX",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=1.2, legend.title.size=0.5,  
            legend.text.size=0.5)
drug_map

```

These three figures will match the maps in Figure 1 of:

Waller LA, Zhu L, Gotway CA, Gorman DM, and Gruenewald PJ (2007)
"Quantifying geographic variations in associations between alcohol
distribution and violence: A comparison of geographically weighted
regression and spatially varying coefficient models". Stochastic
Environmental Research and Risk Assessment.21, 573-588.

Let's use tmap to make a multiple-map single figure.

```{r}
tmap_arrange(violence_map, alc_map, drug_map)
```


***

Now we are ready to fit *geographically weighted regression* (GWR).  GWR uses weighting to fit local regression values to allow the slopes associated with each covariate to change (smoothly) over space.  (See Lecture notes for details).  To estimate the association $\beta(s)$ for a location $s$, GWR uses kernel weights (similar to our intensity estimation for point process data) to weight observations (outcomes and covariates) near $s$ more to give an estimate of $\beta$ associated with location $s$. If you move over, the weights change and the estimated association ($\beta$).   The kernels make sure that if you don't move far, the estimate doesn't change much so we get a smooth surface for $\beta$.

We do this in two steps.  

First, we estimate the bandwidth for the kernel using cross validation (CV).

Next, we use the estimated bandwidth to fit the Poisson GWR model.

***

To get the bandwidth, "ggwr.sel" will test the cross validation score to find a minimum.

***

```{r, warning=FALSE }
### Now to fit Poisson GWR!
# The function 'ggwr' in the 'spgwr' package uses syntax similar to 'glm'
# (like we would use in a standard Poisson regression).
# 'longlat' tells the function that our coordinates are in longitude and
# latitude coordinates.
# 'ggwr.sel' selects the bandwidth for GWR based on the data, the model,
# and cross-validation
# Create distance matrix from centroids

houston.sp <- houston %>%
  as('Spatial')
DM <- gw.dist(dp.locat = coordinates(houston.sp))
head(houston)
houston.bw <- bw.gwr (violence_2 ~ Zl_total + Zl_drug + offset(log(POP2000)),
             data = houston.sp, 
             adaptive=TRUE, 
             dMat=DM)

bgwr <- ggwr.basic(violence_2 ~ Zl_total + Zl_drug + offset(log(POP2000)), 
                      data =houston.sp,
                      family = "poisson",
                      bw = houston.bw, 
                      kernel = "bisquare", 
                      adaptive = TRUE,
                      dMat = DM)
```


***
Now to map the results.

The houston.ggwr object contains an element called SDF which contains the spatial data frame (SDF) of the outcomes.  We will want to map the (spatially varying) intercept and the parameters associated with each of our two covariates.  The notation is tricky since the spatial data frame is within the houston.ggwr object.
 
***

Plot the map of intercept estimates

```{r}
intercept_map <- tm_shape(bgwr$SDF) + 
  tm_fill('Intercept', 
style='quantile', 
palette=subset_colors, 
title='GWR Intercepts') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Intercept",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=0.8, legend.title.size=0.5,  
            legend.text.size=0.5)
intercept_map
```

Plot the map of estimates of association with standardized total alcohol sales.

```{r}
beta_alcohol_map <- tm_shape(bgwr$SDF) + 
  tm_fill('Zl_total', 
style='quantile', 
palette=subset_colors, 
title='GWR beta-hat for standardized total alcohol sales') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Beta Standardized Total Alcohol Sales",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=0.8, legend.title.size=0.5,  
            legend.text.size=0.5)
beta_alcohol_map
```

```{r}
beta_drug_map <- tm_shape(bgwr$SDF) + 
  tm_fill('Zl_drug', 
style='quantile', 
palette=subset_colors, 
title='GWR beta-hat for standardized illegal drug arrests') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Beta Standardized Illegal Drug Arrests",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=0.8, legend.title.size=0.5,  
            legend.text.size=0.5)
beta_drug_map
```

Now make multi-map figure of all three maps

```{r}
tmap_arrange(intercept_map, beta_alcohol_map, beta_drug_map)
```

***

These are very smooth maps.   Let's try it with a smaller bandwidth.

***

```{r, warning=FALSE}
# Try with a smaller, fixed bandwidth (1/4 the size of the other)
# This is closer to what is presented in Figure 5 of Waller et al. 2007

ggwr.bwdiv4 =ggwr.basic(violence_2 ~ Zl_total + Zl_drug + offset(log(POP2000)), 
                      data =houston.sp,
                      family = "poisson",
                      bw = (houston.bw/4), 
                      kernel = "bisquare", 
                      adaptive = TRUE,
                      dMat = DM)
                    
# Now to map the outcomes

# The houston.ggwr object contains an element called SDF which contains
# the spatial data frame (SDF) of the outcomes.
# We will want to map the (spatially varying) intercept and the parameters
# associated with each of our two covariates.
#  The notation is tricky since the spatial data frame is within the
# houston.ggwr object.
# ggwr$SDF@data
intercepts = ggwr.bwdiv4$SDF@data$Intercept
alcohol.effects = ggwr.bwdiv4$SDF@data$Zl_total
drug.effects = ggwr.bwdiv4$SDF@data$Zl_drug
```

Plot the map of intercept estimates

```{r}
intercept_map_bwdiv4 <- tm_shape(ggwr.bwdiv4$SDF) + 
  tm_fill('Intercept', 
style='quantile', 
palette=subset_colors, 
title='GWR Intercepts') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Intercept",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=0.8, legend.title.size=0.5,  
            legend.text.size=0.5)
intercept_map_bwdiv4
```

Plot the map of estimates of association with standardized total alcohol sales.

```{r}
beta_alcohol_map_bwdiv4 <- tm_shape(ggwr.bwdiv4$SDF) + 
  tm_fill('Zl_total', 
style='quantile', 
palette=subset_colors, 
title='GWR beta-hat for standardized total alcohol sales') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Beta Standardized Total Alcohol Sales",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=0.8, legend.title.size=0.5,  
            legend.text.size=0.5)
beta_alcohol_map_bwdiv4
```

Plot the map of estimates of association with standardized illegal drug arrests

```{r}
beta_drug_map_bwdiv4 <- tm_shape(ggwr.bwdiv4$SDF) + 
  tm_fill('Zl_drug', 
style='quantile', 
palette=subset_colors, 
title='GWR beta-hat for standardized illegal drug arrests') + 
  tm_borders(alpha=0.7) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Beta Standardized Illegal Drug Arrests",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=0.8, legend.title.size=0.5,  
            legend.text.size=0.5)
beta_drug_map_bwdiv4
```


Now make multi-map figure of all three maps

```{r}
tmap_arrange(intercept_map_bwdiv4, beta_alcohol_map_bwdiv4, beta_drug_map_bwdiv4)
```

