---
title: "Waller SISMID 2023:  NY Leukemia Spatial Regression with INLA"
author: "Lance A. Waller"
date: "7/18/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## New York Leukemia Data Spatial Poisson Regression with INLA Example

NY Leukemia anlaysis with INLA from https://www.rdocumentation.org/packages/spdep/versions/0.6-15/topics/NY_data
and https://www.r-bloggers.com/spatial-data-analysis-with-inla/ 

```{r }
##Load libraries

library(spdep)
library(INLA)

# Set my working directory (Lance's here for example)
# setwd("~/OneDrive - Emory University/meetings/SISMID.2021/SISMID.2021.Waller.Rcode")

# Read in data (included in the 'foreign' library.)
library(foreign)
nydata <- read.dbf(system.file("etc/misc/nydata.dbf", package="spdep")[1])
```

Now format the data for INLA (this needs to be more clearly documented, but these steps put the right values from the dataset into the variable names needed for INLA)

```{r }
# Format data for INLA
coordinates(nydata) <- c("X", "Y")
nyadjmat <- as.matrix(read.dbf(system.file("etc/misc/nyadjwts.dbf",
                                           package="spdep")[1])[-1])
ID <- as.character(names(read.dbf(system.file("etc/misc/nyadjwts.dbf",
                                              package="spdep")[1]))[-1])
identical(substring(ID, 2, 10), substring(as.character(nydata$AREAKEY), 2, 10))
nyadjlw <- mat2listw(nyadjmat, as.character(nydata$AREAKEY))
listw_NY <- nb2listw(nyadjlw$neighbours, style="B")


```

Now let's calculate the overall rate (total number of cases over total population size).

Next, find the number of cases in each tract if this overall rate were applied to the tract-level population size.

Finally, calculate the standardized mortality ratio, the number observed divided by the number expected (if the overall rate applied equally everywhere).

```{r }
# Calculate overall rate
rate <- sum(nydata$TRACTCAS) / sum(nydata$POP8)

# Set expected number of cases (population * rate)
Expected <- nydata$POP8 * rate

# Calculate standardized mortality ratio (SMR)
SMR <- nydata$TRACTCAS / nydata$Expected

```

Now fit the ICAR model and output the estimates.

```{r }
# Fit ICAR model
# Assuming here that the order of nydata and nyadjmat are the same
ID_new <- seq(nrow(nydata)) # I think INLA needs the IDs to be 1:n
nydata$ID_new <- ID_new
nyadjlw <- mat2listw(nyadjmat, row.names = ID_new)
listw_NY <- nb2listw(nyadjlw$neighbours, style="B")
# FEED THE FOLLOWING INTO INLA'S graph ARGUMENT instead of listw_NY
nyadj_matrix <- as(nb2mat(nyadjlw$neighbours, style = "B"), "Matrix")

m.icar <- inla(trunc(TRACTCAS) ~ 1 + PCTAGE65P + AVGIDIST + PCTOWNHOME +
                 f(ID_new, model = "besag", graph = nyadj_matrix),
               data = as.data.frame(nydata), E = Expected, family ="poisson",
               control.predictor = list(compute = TRUE),
               control.compute = list(dic = TRUE, waic = TRUE))

summary(m.icar)
```
