---
title: 'Waller SISMID 2024 Reading and Mapping Shapefiles:  Alcohol, Drugs, and Crime in Houston'
author: "Lance A. Waller"
date: "7/15/2024"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

***

* **What we have**
An ArcGIS shapefile for census tracts in Houston.
+ A "shapefile" is actually several separate files with the same file name and different extensions.
+ The .dbf file is a dbase database file containing the attribute table
+ The .shp file contains information on the outlines of the tracks.
+ The .shx, .sbn, and .sbx files have the connective tissue to link it all together (sometimes you can have a .prj paper for the map projection).
The following attributes are of interest to us:  Population (census), violent crimes (police), number of illegal drug arrests (police), total alcohol sales (state alcoholic beverage license data)

***

* **What we want**
+ *Choropleth* maps of the violent crime rate and standardized log drug arrests and standardized log alcohol sales.
+ Will need to calculate these variables, choose intervals and colors, and map.

***

* **How we get it**
+ We will use the 'sf' package and the function 'st_read' 
+ We will use the 'RColorBrewer' and 'classInt' libraries to choose colors and intervals.
+ For this example, we will use the basic 'plot' function to make maps.

***

First, load the libraries. The working directory should automatically be set to the correct one if you opened the R project (`2024-SISMID-Spatial-Epi.Rproj`) through RStudio.  

```{r, warning=FALSE}
##Load libraries
#install.packages("pacman")
pacman::p_load(sf, #replaces "maptools", "rgdal" and other deprecated packages 
               tmap, #helps with plotting your map
               RColorBrewer, # creates nice color schemes
               ClassInt,  # finds class intervals for continuous variables
               spgwr, # Adds the geographically weighted regression functions
               here # For constructing filepaths relative to root directory
               )
```
***

Now to read in the shapefile.  The following five files all constitute what is a "shapefile." If you downloaded the repository from GitHub correctly, they should all be in your `data` folder. 

+ HoustonENAR2012final.shp
+ HoustonENAR2012final.dbf
+ HoustonENAR2012final.shx
+ HoustonENAR2012final.sbx
+ HoustonENAR2012final.sbn

```{r}
##Read in shapefile - Houston Census Tracts
houston = st_read(dsn = here("data"),layer = "HoustonENAR2012final")
```
***

## Plotting the map
+ To get the outlines simply plot our 'houston' data object.

***
```{r}
plot(houston)
```
***

Next, we want to make choropleth maps (shading in each tract based on it attribute value).

To do this, we need to decide how many intervals (colors) we want.  I like odd numbers so there is a 'middle' color, and I usually start with quintiles.

'classInt' will assign each tract to the appropriate quintile for a particular attribute.

'RColorBrewer' will assign a color scheme for the quantiles...lots of fascinating work on color choices by Cynthia Brewer, well worth reading and checking the ColorBrewer webpage (https://colorbrewer2.org/)

```{r}

# Plot the outlines (we'll add color below)
plot(houston)

# Define the variable (attribute) to shade tracts by
pop2000 <- houston$POP2000

# Define the number of classes
nclr <- 5  # quintiles
# Use RColorBrewer to choose the colors
plotclr <- brewer.pal(nclr,"BuPu")

class <- classInt::classIntervals(pop2000, nclr, style="quantile")
colcode <- classInt::findColours(class, plotclr)

#Fill in the tracts with the colors
plot(houston['POP2000'])
#Add a title
title(main="Population 2000",
      sub="Quantile (Equal-Frequency) Class Intervals")
```

OK, great, we can map the data.  Let's map the main variables for our analysis.

The data table has a lot of census data and various transformations of the violent crime, alcohol sales, and drug arrest data.  The next section pulls the values we want.

***
```{r}
# Outcome:  Number of violent crimes by tract
hist(houston$violence_2)

# Divide by the 2000 population to get the rate
houston$violence.rate = houston$violence_2/houston$tot_pop

#Summarize key variables 
summary(houston$violence.rate)
summary(houston$violence_2)
summary(houston$tot_pop)

# Covariate 1 (log standardized total alcohol sales)
hist(houston$Zl_total)
# Covariate 2 (log standardized illegal drug arrests)
hist(houston$Zl_drug)
```
***
Now to map the outcome.
***

```{r}
violence_map <- tm_shape(houston) + 
  tm_fill('violence.rate', 
style='quantile', 
palette='BuPu', 
title='Violent Crimes \n Per Capita') + 
  tm_borders(alpha=0.2) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Violent Crime Rate in Houston, TX",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=1.2, legend.title.size=0.8)
violence_map
```
***
Next, map standardized log total alcohol sales.
***

```{r}
alc_map <- tm_shape(houston) + 
  tm_fill('Zl_total', 
style='quantile', 
palette='BuPu', 
title='Standardized Log \n Total Alcohol Sales') + # "\n" moves text to the next line 
  tm_borders(alpha=0.2) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Alcohol Sales in Houston, TX",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=1.2, legend.title.size=0.8)
alc_map
```
***
Finally, plot standardized log illegal drug arrests.
***

```{r}
drug_map <- tm_shape(houston) + 
  tm_fill('Zl_drug', 
style='quantile', 
palette='BuPu', 
title='Standardized Log \n Illegal Drug Arrests') + 
  tm_borders(alpha=0.2) + 
  tm_credits('Quantile (Equal-Frequency) Class Intervals', 
             position=c('RIGHT', 'BOTTOM')) + 
  tm_layout(main.title="Illegal Drug Arrests in Houston, TX",
            inner.margins = c(0.1, 0.1, 0.05, 0.05), 
            main.title.size=1.2, legend.title.size=0.8)
drug_map

```

***

These three figures will match the maps in Figure 1 of:  

Waller LA, Zhu L, Gotway CA, Gorman DM, and
Gruenewald PJ (2007) "Quantifying geographic variations in
associations between alcohol distribution and violence: A
comparison of geographically weighted regression and spatially
varying coefficient models".  *Stochastic
Environmental Research and Risk Assessment*. **21**, 573-588.
***
```{r}
tmap_arrange(violence_map, alc_map, drug_map)
```
